### 2015-03-15 staticanalysis.tex
updates to instrumentation text

@@

\begin{table*}
\small
\begin{center}
\begin{tabular}{lll}
\textbf{Statement}           & \textbf{Logger's Instrumentation}      & \textbf{Checker's Instrumentation} \\
\noalign{\smallskip}\toprule
\textbf{lock}(m);              & \textbf{add\_lock}(CL, m);                        & \textbf{add\_lock}(CL, m); \\
\noalign{\smallskip}\hline\noalign{\smallskip}
\textbf{unlock}(m);          & \textbf{remove\_lock}(CL, m);                  & \textbf{remove\_lock}(CL, m); \\
\noalign{\smallskip}\hline\noalign{\smallskip}
\multirow{2}{*}{x := *ptr;} & LS$\lbrack $ptr$\rbrack :=$ LS$\lbrack $ptr$\rbrack$ $\cap$ CL; & \textbf{assert}(W$\lbrack $ptr$\rbrack \implies$ CL $\cap$ LS$\lbrack $ptr$\rbrack \not= \varnothing$); \\
\noalign{\smallskip}
                                       & \textbf{havoc}(x);                                 & \textbf{havoc}(x); \\
\noalign{\smallskip}\hline\noalign{\smallskip}
\multirow{2}{*}{*ptr := x;} & LS$\lbrack $ptr$\rbrack :=$ LS$\lbrack $ptr$\rbrack$ $\cap$ CL; & \multirow{2}{*}{\textbf{assert}(CL $\cap$ LS$\lbrack $ptr$\rbrack \not= \varnothing$);} \\
\noalign{\smallskip}
                                       & W$\lbrack $ptr$\rbrack := true;$ &
\end{tabular}
\end{center}
\caption{Logger and Checker statements to be instrumented in each pair of entry points. The \texttt{havoc(x)} statement assigns a nondeterministic value to \texttt{x}.}
\label{tab:instrumentation}
\end{table*}

>>>>

Next, \whoop instruments the program with global variables for lockset analysis and race checking per entry point. For example, for entry point \texttt{ep1}, \whoop instruments:

\begin{boogie}
var {:current_lockset} lock$0_in_CLS_$ep1: bool;
var {:lockset} lock$0_in_LS_$M.0_$ep1: bool;
var {:access_checking} WRITTEN_$M.0_$ep1: bool;
var {:access_checking} READ_$M.0_$ep1: bool;
\end{boogie}

\whoop also instruments a global watchdog variable per memory region that is common to all entry points:

\begin{boogie}
const {:watchdog} WATCHED_ACCESS_$M.0: int;
\end{boogie}

The above instrumented variables represent the follows: \texttt{lock\$0\_in\_CLS\_\$ep1} is the current lockset (CLS) of \texttt{ep1}; \texttt{WATCHED\_ACCESS\_\$M.0} is an \emph{unconstrained} constant representing the offset to memory region \texttt{\$M.0} that should be checked for races; \texttt{lock\$0\_in\_LS\_\$M.0\_\$ep1} is the lockset for \texttt{\$M.0}; \texttt{WRITTEN\_\$M.0\_\$ep1} is the write-set for \texttt{\$M.0}; and \texttt{READ\_\$M.0\_\$ep1} is the read-set for \texttt{\$M.0}. Verification involves proving that two entry points cannot race at the watched offset of every memory region. The arbitrary watched offset implies that every offset of each memory region is race-free. Watchdog race checking has been used before for verifying GPU kernels~\cite{bardsley2014engineering}.

\whoop proceeds by instrumenting each entry point call-graph for lockset analysis by replacing each call to a locking or unlocking function with a \whoop-specific lock and unlock functions. This means that Example~\ref{fig:smack_translation} will become:

\begin{boogie}
procedure ep1(dev: int) modifies $M.0; {
  var $p0, $p1, $p2, $p3, $p4, $p5, $p6: int;
  $bb0:
    $p0 := dev;
    $p1 := $pa($p0, 32, 1);
    $p2 := $pa($pa($p1, 0, 24), 12, 1);
    $p3 := $pa($pa($p2, 0, 12), 0, 1);
    call _UPDATE_CLS_$ep1(lock$0, true);
    $p4 := $pa($pa($p1, 0, 24), 4, 1);
    $M.0[$p4] := 4;
    $p5 := $pa($pa($p1, 0, 24), 12, 1);
    $p6 := $pa($pa($p5, 0, 12), 0, 1);
    call _UPDATE_CLS_$ep1(lock$0, false);
    return;
}
\end{boogie}

The \texttt{\_UPDATE\_CLS\_\$ep1()} is a special function that updates the CLS global variable to either true or false, if the lock is held or released by \texttt{ep1} accordingly:

\begin{boogie}
procedure {:inline 1} _UPDATE_CLS_$ep1(lock: int,
    isLocked: bool);
  modifies lock$0_in_CLS_$ep1;

implementation {:inline 1} _UPDATE_CLS_$ep1(
    lock: int, isLocked: bool) {
  _UPDATE:
    lock$0_in_CLS_$ep1 := (if lock == lock$0 then
      isLocked else lock$0_in_CLS_$ep1);
  return;
}
\end{boogie}

Next, \whoop replaces every write and read access in each entry point call-graph with a \whoop-specific function:

\begin{boogie}
procedure ep1(dev: int) modifies $M.0; {
  var $p0, $p1, $p2, $p3, $p4, $p5, $p6: int;
  $bb0:
    \textbf{$p0 := dev;}
    $p1 := $pa($p0, 32, 1);
    $p2 := $pa($pa($p1, 0, 24), 12, 1);
    $p3 := $pa($pa($p2, 0, 12), 0, 1);
    call _UPDATE_CLS_$ep1(lock$0, true);
    $p4 := $pa($pa($p1, 0, 24), 4, 1);
    call _WRITE_LS_$M.0_$ep1($p4);
    $p5 := $pa($pa($p1, 0, 24), 12, 1);
    $p6 := $pa($pa($p5, 0, 12), 0, 1);
    call _UPDATE_CLS_$ep1(lock$0, false);
    return;
}
\end{boogie}

The \texttt{\_WRITE\_LS\_\$M.0\_\$ep1()} is a special function that computes the intersection of the CLS and the lockset of \texttt{\$M.0} (see Section~\ref{bg:lockset}). It also updates the write-set of \texttt{\$M.0} to true, because there was a write access by \texttt{ep1}:

\begin{boogie}
procedure {:inline 1} _WRITE_LS_$M.0_$ep1(
    ptr: int);
  modifies lock$0_in_LS_$M.0_$ep1,
    WRITTEN_$M.0_$ep1;

implementation {:inline 1} _WRITE_LS_$M.0_$ep1(
    ptr: int) {
  _WRITE:
    goto anon1_Then, anon1_Else;

  anon1_Then:
    assume {:partition} WATCHED_ACCESS_$M.0 ==
      ptr && DEVICE_IS_REGISTERED_$ep;
    lock$0_in_LS_$M.0_$ep := lock$0_in_CLS_$ep && lock$0_in_LS_$M.0_$ep;
    WRITTEN_$M.0_$ep := true;
    return;
  anon1_Else:
    assume {:partition} !(WATCHED_ACCESS_$M.0 ==
      ptr && DEVICE_IS_REGISTERED_$ep);
    return;
}
\end{boogie}

The \texttt{\_READ\_LS\_\$M.0\_\$ep1()} updates the lockset of \texttt{\$M.0} and the read-set accordingly for a read access.

To be sound, \whoop returns a nondeterministic value (using the Boogie keyword \texttt{havoc}) for each read access to a shared memory location. This over-approximates any effects from all the unmodeled threads on the shared state, but can and will result into false positives.

%For each array, two Booleans are introduced to record whether a read from or write to the watched offset has occurred. Initially these Booleans are false, and they are reset at each barrier. Thread s sets the ?read? Boolean to true whenever it reads from the watched offset, and similarly for the ?write? Boolean. A race between s and t is reported if thread t reads from the watched offset and the ?write? Boolean is true, or if thread t writes to the watched offset and either the ?read? or ?write? Boolean is true. The non-deterministic choice per array access is eliminated.

@@

### 2015-03-15 translation.tex
updates to instrumentation text

\begin{lstlisting}[caption = Simple networking entry point in C, label = fig:original_program]
static void entrypoint(struct net_device *dev) {
  struct shared *tp = netdev_priv(dev);
  mutex_lock(&tp->wk.mutex);
  tp->field = 4;
  mutex_unlock(&tp->wk.mutex);
}
\end{lstlisting}

@@

\begin{boogie}[caption = Translation to Boogie IVL using SMACK, label = fig:smack_translation]
procedure entrypoint(dev: int) modifies $M.0; {
  var $p0, $p1, $p2, $p3, $p4, $p5, $p6: int;
  $bb0:
    $p0 := dev;
    $p1 := $pa($p0, 32, 1);
    $p2 := $pa($pa($p1, 0, 24), 12, 1);
    $p3 := $pa($pa($p2, 0, 12), 0, 1);
    call mutex_lock($p3);
    $p4 := $pa($pa($p1, 0, 24), 4, 1);
    $M.0[$p4] := 4;
    $p5 := $pa($pa($p1, 0, 24), 12, 1);
    $p6 := $pa($pa($p5, 0, 12), 0, 1);
    call mutex_unlock($p6);
    return;
}
\end{boogie}

>>>>

\begin{lstlisting}[caption = Simple networking entry point in C, label = fig:original_program]
static void ep1(struct net_device *dev) {
  struct shared *tp = netdev_priv(dev);
  mutex_lock(&tp->wk.mutex);
  tp->field = 4;
  mutex_unlock(&tp->wk.mutex);
}
\end{lstlisting}

@@

\begin{boogie}[caption = Translation to Boogie IVL using SMACK, label = fig:smack_translation]
procedure ep1(dev: int) modifies $M.0; {
  var $p0, $p1, $p2, $p3, $p4, $p5, $p6: int;
  $bb0:
    $p0 := dev;
    $p1 := $pa($p0, 32, 1);
    $p2 := $pa($pa($p1, 0, 24), 12, 1);
    $p3 := $pa($pa($p2, 0, 12), 0, 1);
    call mutex_lock($p3);
    $p4 := $pa($pa($p1, 0, 24), 4, 1);
    $M.0[$p4] := 4;
    $p5 := $pa($pa($p1, 0, 24), 12, 1);
    $p6 := $pa($pa($p5, 0, 12), 0, 1);
    call mutex_unlock($p6);
    return;
}
\end{boogie}
